Verified Nanopasses for Compiling Conditionals

### 2025-08-11 paper.tex
check
added a running example

\lstset{%
language=Python,
basicstyle=\ttfamily\small,
keywordstyle=\ttfamily\small,
%morekeywords={match,case,bool,int,let,begin,if,else,closure},
%deletekeywords={locals},
%escapechar=|,
mathescape=true,
columns=flexible,
%moredelim=[is][\color{red}]{~}{~},
showstringspaces=false,
%inputencoding=utf8
literate={→}{{\to}}1,
}

@@

Towards developing a course on verified compilation, I am interested
in adapting verification techniques from CompCert to the compiler
presented in the \emph{Essentials of Compilation}
textbook~\citep{Siek:2023tr,Siek:2023ue}. That book takes an
incremental approach to teaching compilers, that is, it guides the
students through the creation of multiple compilers for a sequence of
languages, each of which adds a language feature to the previous
language. For the language with Boolean expressions and conditionals,
the Explicate Control pass of the compiler uses the one-function per
context approach of Danvy to translate into an x86-flavored register
transfer language.  \emph{Essentials of Compilation} also employs the
nanopass approach, breaking up the compiler into many small passes
that are easier for student to digest~\citep{Sarkar:2004fk}.

This paper presents a proof of correctness in Agda for four nanopasses
that translate a source language with let binding, integer arithmetic,
conditional expressions and Booleans into an x86-flavored register
transfer language. These four passes correspond to the
following three passes in \emph{Essentials of Compilation}:
\begin{enumerate}
\item Remove Complex Operands
\item Explicate Control
\item Select Instructions
\end{enumerate}
The process of proving of correctness of the Explicate Control pass
made it clear to me that it has two distinct responsibilities that
deserve to be separated:
\begin{itemize}
\item Replacing \texttt{let} expressions with assignment statements.
\item Translating Boolean expressions and conditionals into
  \texttt{goto}-based control-flow.
\end{itemize}
(The RTLgen pass of CompCert also handles the above two
responsibilities.)  So in this paper, the Explicate Control pass is
split into two passes, the first named Lift Locals and the second that
keeps the name Explicate Control.
\begin{enumerate}
\item Remove Complex Operands
\item Lift Locals
\item Explicate Control
\item Select Instructions
\end{enumerate}

@@

In this section we describe how we formalize the syntax and semantics
of the source language and four intermediate languages in Agda:
\begin{enumerate}
\item Integers, Booleans, \texttt{let}, and \texttt{if}
\item Monadic Normal Form
\item Imperative with Structured Control Flow
\item A C-like language with \emph{goto}
\item X86 assembly but with variables
\end{enumerate}
These languages and there semantics are quite standard. However, there
are many other intermediate languages in the literature to choose
from, so why these languages, and why in this particular order?  The
answer is that they happen to represent the images of the four
nanopasses.  So then the question becomes, why do we choose to order
the nanopasses in this way? The main thing to consider is that some
passes make later passes easier because they remove a language feature
that would have been complicated to handle. For example, the Lift
Locals pass removes \texttt{let}, and I found \texttt{let} difficult
to deal with in the proof of correctness for Explicate Control
(\texttt{let} is relatively easy to handle in the code for Explicate
Control). The other thing to consider is that some compiler passes use
certain language features in their generated code, so the passes that
remove those language features better come later in the ordering.  For
example, the Closure Conversion pass (not discussed here) generates
code with \texttt{let}, so Closure Conversion comes before Lift
Locals.

One of the perennial challenges of formalizing semantics in a proof
assistant is that one needs to choose a representation of variables.
%% There are many
%% options, and I've tried most of them. But just like a game of
%% whac-a-mole, the savings achieved in one place are lost in others.
Of the many options, my current favorite is good old de Bruijn
indices~\citep{Bruijn:1972kx}.  Thus, identifiers \lstinline{Id} are
represented as natural numbers.

@@

\begin{lstlisting}
compile e = (select-inst ∘ explicate ∘ lift-locals ∘ rco) e
\end{lstlisting}

@@

We implement the Remove Complex Operands pass with the following
\lstinline{rco} function in Agda. This pass is responsible for making
sure that the arguments of primitive operations, such as
\lstinline{Sub} and \lstinline{Eq}, are atomic expression (variables
or literals).

\begin{lstlisting}
rco : Exp $→$ Mon
\end{lstlisting}

Most of the cases of \lstinline{rco} are trivial, but let us focus on
the case for subtraction. We shall start with a straw-man
translation. We recursively compile the subexpressions $e_1$ and
$e_2$. The results might be complex (not atomic), so we
\lstinline{let} binding them both and replace them in the subtraction
with variables (de Bruijn indices 0 and 1). Note that we must shift
the variables in $e_2$ up by one because we inserted a new
\lstinline{let} above $e_2$. (In general, \lstinline{shift-mon m c}
increases by one each variable in $m$ that is greater or equal to the
cutoff $c$.)

@@

\begin{lstlisting}
nth-++-shift-var : ($ρ₁$ $ρ₂$ : List A)(v : A)(x : Id)
  $→$ nth ($ρ₁$ ++ v $∷$ $ρ₂$) (shift-var x (length $ρ₁$))
    $≡$ nth ($ρ₁$ ++ $ρ₂$) x

@@

The Lift Locals pass is responsible for replacing all of the
\lstinline{let} expressions with assignment statements.  After this
pass, variables are still represented by de Bruijn indices but they
are all bound by the \lstinline{Program} AST node.

The \lstinline{lift-locals} function delegates to the recursive
function \lstinline{lift-locals-mon}, where all the work is done.
That function returns a number $n$ that counts the number of
\lstinline{let} bindings in the expression, and the translated
expression.

@@

@@

\noindent Here are the signatures of the four functions that make up
the Explicate Control pass. Next we discuss select cases of these
functions.

\begin{lstlisting}
explicate : Imp-Prog $→$ CProg
explicate-tail : Imp-Exp $→$ Build CStmt
explicate-assign : Id $→$ Imp-Exp $→$ CStmt $→$ Build CStmt
explicate-pred : Imp-Exp $→$ CStmt $→$ CStmt $→$ Build CStmt
\end{lstlisting}

@@

Throughout the Explicate Control pass we added code to the
control-flow graph whenever we needed to avoid code
duplication. However, some further care is needed to avoid adding (1)
trivial nodes that only contain a \lstinline{Goto} and (2) nodes that
are never the target of a jump (i.e., dead code). The
\lstinline{add-node} function can easily handle (1) by inspecting its
argument and if it is a \lstinline{Goto}, just return its label
instead of adding a new node to the CFG.  Avoiding dead code (2) is
more challenging, either requiring a post-processing of the CFG or use
lazy evaluation to delay the addition of nodes to the CFG to the point
where they are definitely needed~\citep{Siek:2023tr}.

@@

The case for conditional expressions also two lemmas about the
monotonicity of the control flow graph.  The induction hypothesis for
\lstinline{e$₂$} says that the generated \lstinline{c$₂$} evaluates
using the CFG that was produced by the compilation of
\lstinline{e$₂$}, but to satisfy the induction hypothesis
for \lstinline{e$₁$} we need \lstinline{c$₂$} evaluated using the CFG
that was produced by the compilation of \lstinline{e$₃$}.
We define the following notation for
expression that one CFG is a prefix of another CFG:
\begin{lstlisting}
G$₁$ $↝$ G$₂$ = $∃$[ G ] G$₁$ ++ G $≡$ G$₂$
\end{lstlisting}
\noindent We prove that the output CFG of \lstinline{explicate-tail} is
an extension of its input CFG.
\begin{lstlisting}
explicate-tail-CFG : $∀${e}{G$₁$ G$₂$}{c}
  $→$ explicate-tail e G$₁$ $≡$ (c , G$₂$)
  $→$ G$₁$ $↝$ G$₂$
\end{lstlisting}
\noindent And we prove that the big-step semantics is monotonic with
respect to the $↝$ relation on control flow graphs.
\begin{lstlisting}
eval-CFG : $∀${c}{$ρ$ $ρ′$}{B$₁$ B$₂$}{s s$′$}{v}
  → B$₁$ $↝$ B$₂$
  → (s , $ρ$) , B$₁$ $⊢ᶜ$ c $⇓$ v $⊣$ (s$′$ , $ρ′$)
  → (s , $ρ$) , B$₂$ $⊢ᶜ$ c $⇓$ v $⊣$ (s$′$ , $ρ′$)
\end{lstlisting}

@@

\noindent The case for conditional expressions needs another lemma
regarding control flow graphs. Similar to \lstinline{explicate-tail},
we need a lemma which states that the output CFG of
\lstinline{explicate-assign} is an extension of its input CFG.

\begin{lstlisting}
explicate-assign-CFG : $∀${x}{m}{c c'}{G$₁$ G$₂$}
  → explicate-assign x m c G$₁$ $≡$ (c' , G$₂$)
  → G$₁$ $↝$ G$₂$
\end{lstlisting}

@@

The responsibility of the Select Instructions pass is to convert the
assignment statements and primitive operations of the C-like language
into x86 instructions. The minor challenge is that many x86
instructions take just two arguments and have an update-in-place
semantics.

@@

\begin{lstlisting}
select-stmt (IfEq a₁ a₂ thn els) =
  CmpQ (to-arg a₁) (to-arg a₂) ∷
  JmpEq thn ∷
  Jmp els ∷ []
\end{lstlisting}

\noindent Finally, we translate \lstinline{Goto} into a \lstinline{Jmp}.
\begin{lstlisting}
select-stmt (Goto l) = Jmp l ∷ [] 
\end{lstlisting}

@@

\begin{lstlisting}
select-inst-correct : $∀${p}{s}{v}
  → interp-prog p s v
  → interp-x86-var (select-inst p) s v
\end{lstlisting}

>>>>

\lstset{%
language=Python,
basicstyle=\ttfamily\small,
keywordstyle=\ttfamily\small,
%morekeywords={match,case,bool,int,let,begin,if,else,closure},
%deletekeywords={locals},
escapechar=|,
mathescape=true,
columns=flexible,
%moredelim=[is][\color{red}]{~}{~},
showstringspaces=false,
%inputencoding=utf8
literate={→}{{\to}}1,
}

@@

Towards developing a course on verified compilation, I am interested
in adapting verification techniques from CompCert to the compiler
presented in the \emph{Essentials of Compilation}
textbook~\citep{Siek:2023tr,Siek:2023ue}. That book takes an
incremental approach to teaching compilers, that is, it guides the
students through the creation of multiple compilers for a sequence of
languages, each of which adds a language feature to the previous
language. For the language with Boolean expressions and conditionals,
the Explicate Control pass of the compiler uses the one-function per
context approach of Danvy to translate into an x86-flavored register
transfer language.  \emph{Essentials of Compilation} also employs the
nanopass approach, breaking up the compiler into many small passes
that are easier for student to digest~\citep{Sarkar:2004fk}.  There
are many other pedogogical treatments of
compilation~\citep{Appel:2003fk,Sumii:2005aa,Aho:2006wb,Cooper:2011aa,Grune:2012aa};
what makes the \emph{Essentials of Compilation} unique is that it is
the first textbook to apply the incremental approach.  However,
\emph{Essentials of Compilation} does not include material about the
formal verification of compilers; hence my present interest.

This paper presents a proof of correctness in Agda for four nanopasses
that translate a source language with let binding, integer arithmetic,
conditional expressions and Booleans into an x86-flavored register
transfer language. The Agda code is available in the supplementary
material. The four compiler passes presented here correspond to the
following three passes in \emph{Essentials of Compilation}:
\begin{enumerate}
\item Remove Complex Operands
\item Explicate Control
\item Select Instructions
\end{enumerate}
The process of proving of correctness of the Explicate Control pass
made it clear to me that it has two distinct responsibilities that
deserve to be separated:
\begin{itemize}
\item Replacing \texttt{let} expressions with assignment statements.
\item Translating Boolean expressions and conditionals into
  \texttt{goto}-based control-flow.
\end{itemize}
(The RTLgen pass of CompCert also handles the above two
responsibilities.)  So in this paper, the Explicate Control pass is
split into two passes, the first named Lift Locals and the second that
keeps the name Explicate Control.
\begin{enumerate}
\item Remove Complex Operands
\item Lift Locals
\item Explicate Control
\item Select Instructions
\end{enumerate}
In compilers that do not take the nanopass approach, unrelated
responsibilities are sometimes combined into the same pass.  For
example, \citet{Sumii:2005aa} combines the first three of the above
passes into a single pass.

@@

In this section we describe how we formalize the syntax and semantics
of the source language and four intermediate languages in Agda:
\begin{enumerate}
\item Integers, Booleans, \texttt{let}, and \texttt{if}
\item Monadic Normal Form
\item Imperative with Structured Control Flow
\item A C-like language with \emph{goto}
\item X86 assembly but with variables
\end{enumerate}
These languages and their semantics are quite standard. However, there
are many other intermediate languages in the literature to choose
from~\citep{Flanagan:1993cg,Hatcliff:1994vn,Kelsey:1995uq,Birkedal:1996aa},
so why these languages, and why in this particular order?  The answer
is that they happen to represent the images of the four nanopasses.
So then the next question is: why do we order the nanopasses in this
way? The main thing to consider is that some passes make later passes
easier because they remove a language feature that would have been
complicated to handle. For example, the Lift Locals pass removes
\texttt{let}, and I found \texttt{let} difficult to deal with in the
proof of correctness for Explicate Control (\texttt{let} is relatively
easy to handle in the code for Explicate Control). The other thing to
consider is that some compiler passes use certain language features in
their generated code, so the passes that remove those language
features better come later in the ordering.  For example, the Closure
Conversion pass (not discussed here) generates code with \texttt{let},
so Closure Conversion comes before Lift Locals. Finally, many design
choices in \emph{Essentials of Compilation} have been inherited from
\citet{Dybvig:2010aa}

One of the perennial challenges of formalizing semantics in a proof
assistant is that one needs to choose a representation of variables.
Of the many options, my current favorite is good old de Bruijn
indices~\citep{Bruijn:1972kx}.  Thus, identifiers \lstinline{Id} are
represented as natural numbers.

@@

\begin{lstlisting}[basicstyle=\ttfamily\footnotesize]
compile e = (select-inst $∘$ explicate $∘$ lift-locals $∘$ rco) e
\end{lstlisting}

To illustrate each pass, we use the source program in
figure~\ref{fig:running-example-source} as a running example,
presented in a concrete syntax with variables as names instead of de
Bruijn indices to enhance readability.  Given the inputs
\lstinline{5}, \lstinline{10}, then \lstinline{0}, the result of this
program is \lstinline{3}.

\begin{figure}[tp]
  \hrule
\begin{lstlisting}
    let x = read in
    let y = read in
    if (if x == 0
        then true
        else read == 1)
    then y - 1
    else (y - 2) - x
\end{lstlisting}
  \hrule
\caption{An example source program.}
\label{fig:running-example-source}
\end{figure}

@@

The Remove Complex Operands pass is responsible for making sure that
the arguments of primitive operations, such as \lstinline{Sub} and
\lstinline{Eq}, are atomic expression (variables or literals). It does
so by insertion \lstinline{let} bindings for new temporary variables.
In the running example (figure~\ref{fig:running-example-monadic}), the
subexpressions \lstinline{read} and \lstinline{y - 2} are both
arguments of primitive operations, so the Remove Complex Operands pass
binds them to the temporary variables \lstinline{t1} and
\lstinline{t2}, respectively.

\begin{figure}[tp]
  \hrule
\begin{lstlisting}
    let x = read in
    let y = read in
    if (if x == 0
        then true
        else (let t1 = read in t1 == 1))
    then (y - 1)
    else (let t2 = y - 2 in t2 - x)
\end{lstlisting}
  \hrule
\caption{Example after Remove Complex Operands, in
  monadic normal form.}
\label{fig:running-example-monadic}
\end{figure}

We implement the Remove Complex Operands pass with the following
\lstinline{rco} function in Agda.
\begin{lstlisting}
rco : Exp $→$ Mon
\end{lstlisting}

Most of the cases of \lstinline{rco} are trivial, but let us focus on
the case for subtraction. We shall start with a straw-man
translation. We recursively compile the subexpressions $e_1$ and
$e_2$. The results might be complex (not atomic), so we
\lstinline{let}-bind them both and replace them in the subtraction
with variables (de Bruijn indices 0 and 1). Note that we must shift
the variables in $e_2$ up by one because we inserted a new
\lstinline{let} above $e_2$. (In general, \lstinline{shift-mon m c}
increases by one each variable in $m$ that is greater or equal to the
cutoff $c$.)

@@

\begin{lstlisting}[basicstyle=\ttfamily\footnotesize]
nth-++-shift-var : ($ρ₁$ $ρ₂$ : List A)(v : A)(x : Id)
  $→$ nth ($ρ₁$ ++ v $∷$ $ρ₂$) (shift-var x (length $ρ₁$))
    $≡$ nth ($ρ₁$ ++ $ρ₂$) x

@@

The Lift Locals pass is responsible for replacing \lstinline{let}
expressions with assignment statements. In the output language,
variables are declared at the top of the program.  Here is the result
of Lift Locals on the running example. The output of Lift Locals on
the running example is shown in
figure~\ref{fig:running-example-structured-imperative}.

\begin{figure}[tp]
  \hrule
\begin{lstlisting}
    var x, y, t1, t2;
    x := read;
    y := read;
    if (if x == 0
        then true
        else (t1 := read; t1 == 1)
    then (y - 1)
    else (t2 := y - 2; t2 - x)
\end{lstlisting}
  \hrule
  \caption{Example after Lift Locals, in an imperative IR.}
  \label{fig:running-example-structured-imperative}
\end{figure}
The \lstinline{lift-locals} function delegates to the recursive
function \lstinline{lift-locals-mon}, where all the work is done.
That function returns a number $n$ that counts the number of
\lstinline{let} bindings in the expression, and the translated
expression.

@@

The output of Explicate Control on the running example is shown in
figure~\ref{fig:running-example-goto}. Note that there are two
\lstinline{goto} statements that target \lstinline{block_2}, so indeed
the control-flow graph of this program is a directed acyclic graph
(DAG) and not a tree. The DAG representation enables sharing so the
pass can avoid the duplication of code. We discuss the trivial blocks
4 and 5 at the end of this section.

\begin{figure}[tp]
  \hrule
\begin{lstlisting}
    var x, y, t1, t2;
    block_0:
       x := read;
       y := read;
       if (x == 0) goto block_1; else goto block_3;
    block_1:
        goto block_2;
    block_2:
        return y - 1;
    block_3:
       t1 := read;
       if (t1 == 1) goto block_4; else goto block_5;
    block_4:
        goto block_2;
    block_5:
        goto block_6;
    block_6:
        t2 := y - 2;
        return t2 - x;
\end{lstlisting}
  \hrule
  \caption{Example after Explicate Control, with control flow
    made explicit with goto.}
\label{fig:running-example-goto}
\end{figure}

@@

\noindent The following four functions make up the Explicate Control
pass. Next we discuss select cases of these functions.

\begin{lstlisting}[basicstyle=\ttfamily\footnotesize]
explicate : Imp-Prog $→$ CProg
explicate-tail : Imp-Exp $→$ Build CStmt
explicate-assign : Id $→$ Imp-Exp $→$ CStmt $→$ Build CStmt
explicate-pred : Imp-Exp $→$ CStmt $→$ CStmt $→$ Build CStmt
\end{lstlisting}

@@

Throughout the Explicate Control pass we added code to the
control-flow graph whenever we needed to avoid code
duplication. However, some further care is needed to avoid adding (1)
trivial nodes that only contain a \lstinline{Goto} and (2) nodes that
are never the target of a jump (i.e., dead code). The
\lstinline{add-node} function can easily handle (1) by inspecting its
argument and, if it is a \lstinline{Goto}, return its label instead of
adding a new node to the CFG. This optimization is known as
compressing corridor transitions~\citep{Danvy:2009aa}.  Avoiding dead
code (2) is more challenging, either post-process the CFG or use lazy
evaluation to delay the addition of nodes to the CFG to the time when
they are definitely needed~\citep{Siek:2023tr}.

@@

The case for conditional expressions also two lemmas about the
monotonicity of the control flow graph.  The induction hypothesis for
\lstinline{e$₂$} says that the generated \lstinline{c$₂$} evaluates
using the CFG that was produced by the compilation of
\lstinline{e$₂$}, but to satisfy the induction hypothesis
for \lstinline{e$₁$} we need \lstinline{c$₂$} evaluated using the CFG
that was produced by the compilation of \lstinline{e$₃$}.
We define the following notation for
expression that one CFG is a prefix of another CFG:
\begin{lstlisting}
G$₁$ $↝$ G$₂$ = $∃$[ G ] G$₁$ ++ G $≡$ G$₂$
\end{lstlisting}
\noindent We prove that the output CFG of \lstinline{explicate-tail} is
an extension of its input CFG.
\begin{lstlisting}
explicate-tail-CFG : $∀${e}{G$₁$ G$₂$}{c}
  $→$ explicate-tail e G$₁$ $≡$ (c , G$₂$)
  $→$ G$₁$ $↝$ G$₂$
\end{lstlisting}
\noindent And we prove that the big-step semantics is monotonic with
respect to the $↝$ relation on control flow graphs.
\begin{lstlisting}
eval-CFG : $∀${c}{$ρ$ $ρ′$}{B$₁$ B$₂$}{s s$′$}{v}
  $→$ B$₁$ $↝$ B$₂$
  $→$ (s , $ρ$) , B$₁$ $⊢ᶜ$ c $⇓$ v $⊣$ (s$′$ , $ρ′$)
  $→$ (s , $ρ$) , B$₂$ $⊢ᶜ$ c $⇓$ v $⊣$ (s$′$ , $ρ′$)
\end{lstlisting}

@@

\noindent The case for conditional expressions needs another lemma
regarding control flow graphs. Similar to \lstinline{explicate-tail},
we need a lemma which states that the output CFG is an extension of
its input CFG.

\begin{lstlisting}
explicate-assign-CFG : $∀${x}{m}{c c'}{G$₁$ G$₂$}
  $→$ explicate-assign x m c G$₁$ $≡$ (c' , G$₂$)
  $→$ G$₁$ $↝$ G$₂$
\end{lstlisting}

@@

The responsibility of the Select Instructions pass is to convert the
assignment statements and primitive operations of the C-like language
into x86 instructions. The minor challenge is that many x86
instructions take just two arguments and have an update-in-place
semantics. The output of Select Instructions on the running example is
shown in figure~\ref{fig:running-example-x86}.\footnote{The redundant
\texttt{movq} instructions are removed by the Patch Instructions pass
in \emph{Essentials of Compilation}, which comes after Select
Instructions and Register Allocation.}

\begin{figure}[tp]
  \hrule
\begin{lstlisting}[mathescape=false]
    var x, y, t1, t2
    block_0:
       callq read_int
       movq %rax, x
       callq read_int
       movq %rax, y
       cmpq x, $0
       je block_1
       jmp block_3
    block_1:
        jmp block_2
    block_2:
        movq y, %rax
        subq $1, %rax
        movq %rax, %rax
    block_3:
        callq read_int
        movq %rax, t1
        cmpq t1, $1
        je block_4
        jmp block_5
    block_4:
        jmp block_2
    block_5:
        jmp block_6
    block_6:
        movq y, %rax
        subq $2, %rax
        movq %rax, t2
        movq t2, %rax
        subq x, %rax
        movq %rax, %rax
\end{lstlisting}
  \hrule
  \caption{Example after Select Instructions, in the x86 assembly language
    augmented with variables.}
\label{fig:running-example-x86}
\end{figure}

@@

\begin{lstlisting}
select-stmt (IfEq a$_1$ a$_2$ thn els) =
  CmpQ (to-arg a$_1$) (to-arg a$_2$) $∷$
  JmpEq thn $∷$
  Jmp els $∷$ []
\end{lstlisting}

\noindent Finally, we translate \lstinline{Goto} into a \lstinline{Jmp}.
\begin{lstlisting}
select-stmt (Goto l) = Jmp l $∷$ [] 
\end{lstlisting}

@@

\begin{lstlisting}
select-inst-correct : $∀${p}{s}{v}
  $→$ interp-prog p s v
  $→$ interp-x86-var (select-inst p) s v
\end{lstlisting}
